---
output:
  html_document:
    theme: readable
resource_files:
- index.html
runtime: shiny
---

#Degeneracy, instability, and uninterpretability... Oh my!

```{r libraries, message=FALSE, warning=FALSE, echo=FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(shiny)
library(gridExtra)

opts_chunk$set(echo=FALSE, message=FALSE, warnings=FALSE)
theme_set(theme_bw(base_family="serif"))

source("functions.R")
```

```{r}
load("grid_measures.RData")

top_plots <- list(p1 = three_ways %>%
    ggplot() +
    geom_tile(aes(x = r1, y = r2, fill = frac_degen)) +
    geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", bins = 8) +
    geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", breaks = .05, size = 1.5) +
    #geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
    scale_fill_gradient("Fraction near-degenerate", low = "yellow", high = "red") +
    facet_grid(Visibles~Hiddens) +
    xlab(expression(group("||", theta[main], "||"))) +
    ylab(expression(group("||", theta[interaction], "||"))) +
    theme(aspect.ratio = 1) +
    theme(legend.position = "bottom") +
    ggtitle("Near-degeneracy"),
    p2 =  three_ways %>%
    ggplot() +
    geom_tile(aes(x = r1, y = r2, fill = mean_abs_diff)) +
    geom_contour(aes(x = r1, y = r2, z = mean_abs_diff), colour = "black", bins = 8) +
    #geom_contour(aes(x = r1, y = r2, z = mean_abs_diff), colour = "black", breaks = .05, size = 1.5) +
    #geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
    scale_fill_gradient(expression(group("|", E(group("", bold(X), "|")*bold(theta)) - E(group("", bold(X), "|")*plain(independence)), "|")), low = "yellow", high = "red") +
    facet_grid(Visibles~Hiddens) +
    xlab(expression(group("||", theta[main], "||"))) +
    ylab(expression(group("||", theta[interaction], "||"))) +
    theme(aspect.ratio = 1) +
    theme(legend.position = "bottom") +
    ggtitle("Uninterpretability"),
    p3 = three_ways %>%
    ggplot() +
    geom_tile(aes(x = r1, y = r2, fill = mean_max_q)) +
    geom_contour(aes(x = r1, y = r2, z = mean_max_q), colour = "black", bins = 8) +
    #geom_contour(aes(x = r1, y = r2, z = mean_max_q), colour = "black", breaks = .05, size = 1.5) +
    #geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
    scale_fill_gradient(expression(frac(max(Q), H + V)), low = "yellow", high = "red") +
    facet_grid(Visibles~Hiddens) +
    xlab(expression(group("||", theta[main], "||"))) +
    ylab(expression(group("||", theta[interaction], "||"))) +
    theme(aspect.ratio = 1) +
    theme(legend.position = "bottom") +
    ggtitle("Instability")
)

inputPanel(
  selectizeInput("variable", "Select radius to fix:", 
                 choices = list("Main effects" = "r1", "Interaction effects" = "r2"), 
                 multiple = FALSE,
                 selected = "r1"),
  numericInput("slice", "Fix radius at:",
               min = min(three_ways$r1), 
               max = max(three_ways$r1), 
               value = min(three_ways$r1), step = .1)
)

renderPlot({

  if(input$variable == "r1") {
    p1 <- top_plots$p1 + geom_vline(aes(xintercept = input$slice))
    p2 <- top_plots$p2 + geom_vline(aes(xintercept = input$slice))
    p3 <- top_plots$p3 + geom_vline(aes(xintercept = input$slice))
  } else {
    p1 <- top_plots$p1 + geom_hline(aes(yintercept = input$slice))
    p2 <- top_plots$p2 + geom_hline(aes(yintercept = input$slice))
    p3 <- top_plots$p3 + geom_hline(aes(yintercept = input$slice))
  }
  
  grid.arrange(p1, p3, p2, nrow = 1)
})

filter_dat <- reactive({
    slice <- unique(three_ways[which.min(abs(three_ways[, input$variable] %>% data.matrix() - input$slice)), input$variable])[1] %>% 
    as.numeric()
    
    three_ways[three_ways[,input$variable] == slice, ]
})

renderPlot({
  axes <- c("r1", "r2")
  variable <- axes[axes != input$variable]
  label <- ifelse(variable == "r1", 
                  expression(group("||", theta[main], "||")), 
                  expression(group("||", theta[interaction], "||")))
  
  dat <- filter_dat()
  
   dat %>%
    ggplot() +
    geom_line(aes_string(variable, "frac_degen")) +
    facet_grid(Visibles~Hiddens) +
    xlab(label) +
    ylab("Fraction Near-degenerate") +
    theme(aspect.ratio = 1) +
    ggtitle("Near-degeneracy") -> p1
  
  dat %>%
    ggplot() +
    geom_line(aes_string(variable, "mean_abs_diff")) +
    facet_grid(Visibles~Hiddens) +
    xlab(label) +
    ylab(expression(group("|", E(group("", bold(X), "|")*bold(theta)) - E(group("", bold(X), "|")*plain(independence)), "|"))) +
    theme(aspect.ratio = 1) +
    ggtitle("Uninterpretability") -> p2
  
  dat %>%
    ggplot() +
    geom_line(aes_string(variable, "mean_max_q")) +
    facet_grid(Visibles~Hiddens) +
    xlab(label) +
    ylab(expression(frac(max(Q), H + V))) +
    theme(aspect.ratio = 1) +
    ggtitle("Instability") -> p3
  
  grid.arrange(p1, p3, p2, nrow = 1)
})

```

The highly flexible nature of the RBM ($H + V + HV$ parameters) makes the following characteristics of model impropriety of particular concern. 

 - **Near-degeneracy.** Occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space by the model. If random variables in the neg-potential function $Q(\cdot)$, having support set $\mathcal{S}$, have a collective mean $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{S}$.
 - **Instability.** Small changes in natural parameters result in large changes in probability masses, excessive sensitivity. If for any $C > 0$ there exists $N_C > 0$ such that 
$\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] > CN$ for all $N > N_C$, where $Q(\cdot)$ is the neg-potential function of the model.
 - **Uninterpretability.** Due to the existence of dependence, marginal mean-structure no longer maintained. If the magnitude of the difference between model expectations and expectations under independence (dependence parameters of zero), $\left|\text{E}(\boldsymbol X | \boldsymbol \theta) - \text{E}(\boldsymbol X | \emptyset ) \right|$, is large.
 
 
 